Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-11-13 manuscript/review_response_R1.Rmd

**Reviewer: 3**

*Basic reporting*

English is high quality with very few typos, though a few locations in text could use rephrasing to increase clarity. Some restructuring of the introduction is suggested to provide a more logical and engaging presentation of the topic. Whenever I had a question about a topic the authors were referring to, or a method they were using, their citations were sufficient for me to conduct further investigation. More discussion of prior work into the role of dominant species in determining ecosystem function may be warranted.

\response{These are all good suggestions. Reviewer 2 agreed that the paper would benefit from a broader introduction, which we have added (lines xx-xx). We have also added a brief reference to the idea that domoninant species play a large role in determining ecosystem functioning (lines xx-xx).}

@@

This is a good article that yields insight into how experimental work can allow us to test whether historic precipitation/biomass relationships will told up to potential future changes in the precipitation regime. Overall, this is a very strong study and paper; it will make an excellent contribution to the scientific literature. However, there are some small issues that should be resolved first. There are some organizational issues in the Introduction; particularly, it seems the motivation for the study is buried deep within the text instead of in the first paragraph. Revising this section would make the paper a much more compelling read. Additionally, there were a number of places in the methods where I felt that additional clarity was needed so that others could replicate their study. Below I give line-by-line suggestions on these points and other minor editing suggestions. This reviewer feels that this paper should be accepted with minor revisions.

\response{Thanks for these thoughtful comments. We address your specific points below.}

General comments to authors

Lines 31-32: sentence beginning with "However, ANPP response…" seems fairly vague and weakly worded, consider revising to increase clarity.

\response{We have revised this sentence to read, "However, all estimates for the effect of soil moisture on ANPP overlapped zero, indicating the relationship is weak and uncertain regardless of treatment." (lines xx-xx).}

@@

Introduction:

While the abstract opens with a strong justification tying the paper into shifting precipitation patterns as a result of climate change and the need to develop a way to predict how communities might change with long-term exposure, the introduction opens only with a statement saying that there is a relationship between soil moisture and ANPP. This seems incredibly weak and uncompelling. The authors should consider restructuring their introduction so that the justification for their study is at the beginning. They have several other paragraphs in the paper that could do nicely as an opening paragraph. The opening paragraph of the discussion Lines 262-269 would make for an excellent opening. Or, with a little retooling the second paragraph of the introduction (lines 53-64) would make for a better statement of research motivation than the current opening.

\response{Fair point, and one that echoes Reviewer 2's critique of our introduction. We have added a new opening paragraph that we hope better sets the stage (lines xx-xx).}

### 2017-11-13 manuscript/usses_water_ms.Rmd
more minor revisions

  **Results.** Ecosystem functional response, defined as the relationship between soil moisture and ANPP was similar among drought and control treatments, but the irrigation treatment had a lower slope than the control treatment. However, ANPP response to available soil moisture was weak and uncertain regardless of treatment, with all slopes overlapping zero. There was also large spatial variation in ANPP within-years. Plant community composition was remarkably stable over the course of the experiment and did not differ among treatments.

@@

We monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors around the edges of each 1x1 m census plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings taken at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06-06-2012, 04-29-2015, 05-07-2015, 06-09-2015, and 05-10-2016\footnote{Dates formatted as: mm-dd-yyyy.} using a handheld EC-5 sensor.

@@

We first averaged the observed soil moisture for each day and within each plot.
Then we standardized the averages within each plot group by subtracting the average ambient soil moisture in that plot group and dividing by the standard deviation of the ambient soil moisture in that plot group.
We then found the difference between the standardized ambient soil moisture and the standardized drought and irrigation soil moisture within each plot group.
These transformations ensured that the treatment effects in each plot were appropriately scaled by the local ambient conditions within each plot group.

@@

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS (Moderate Resolution Imaging Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer) bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 1).

>>>>

  **Results.** Ecosystem functional response, defined as the relationship between soil moisture and ANPP was similar among drought and control treatments, but the irrigation treatment had a lower slope than the control treatment. However, all estimates for the effect of soil moisture on ANPP overlapped zero, indicating the relationship is weak and uncertain regardless of treatment. There was also large spatial variation in ANPP within-years, which contributes to the uncertainty of the soil moisture effect. Plant community composition was remarkably stable over the course of the experiment and did not differ among treatments.

@@

We monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors around the edges of each 1x1 m census plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings \new{taken with a handheld EC-5 sensor} at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06-06-2012, 04-29-2015, 05-07-2015, 06-09-2015, and 05-10-2016\footnote{Dates formatted as: mm-dd-yyyy.}.

@@

We first averaged the observed soil moisture for each day (*d*) and plot (*i*), $x_{i,d}$.
\new{Then we standardized the average of each plot within a plot group ($g$) by subtracting the mean ambient soil moisture of the plot group for the day ($\overline{x}_{g,d} = \sum^{i=1}_G x_{i,d} / n$) from each plot's daily average and dividing by the standard deviation of the ambient soil moisture in that plot group ($\sigma_{g,d}$): $x^*_{i,d} = \frac{x_{i,d} - \overline{x}_{g,d}}{\sigma_{g,d}}$.}
We then found the difference between the standardized ambient soil moisture and the standardized drought and irrigation soil moisture within each plot group.
These transformations ensured that the treatment effects in each plot were appropriately scaled by the local ambient conditions within each plot group.

@@

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS (Moderate Resolution Imaging Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer) bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 1).
\new{Our analysis, described below, can be done using NDVI directly, but we focus on ANPP to more easily relate our findings to previous work and because our results do not change if we analyze NDVI directly (Appendix 2).}

### 2017-11-13 manuscript/usses_water_ms.tex
more minor revisions

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, ANPP response to available soil
moisture was weak and uncertain regardless of treatment, with all slopes
overlapping zero. There was also large spatial variation in ANPP
within-years. Plant community composition was remarkably stable over the
course of the experiment and did not differ among treatments.

@@

We first averaged the observed soil moisture for each day and within
each plot. Then we standardized the averages within each plot group by
subtracting the average ambient soil moisture in that plot group and
dividing by the standard deviation of the ambient soil moisture in that
plot group. We then found the difference between the standardized
ambient soil moisture and the standardized drought and irrigation soil
moisture within each plot group. These transformations ensured that the
treatment effects in each plot were appropriately scaled by the local
ambient conditions within each plot group.

@@

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS (Moderate Resolution Imaging
Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer)
bands for NDVI. We regressed the greenness index against the dry biomass
weight from the ten calibration plots to convert the greenness index to
ANPP. We fit regressions to a MODIS-based index and an AVHRR-based index
for each year and retained the regression with the better fit based on
\(R^2\) values. We then predicted ANPP using the best regression
equation for each year (Appendix 1).

@@

Our results suggest the species in our focal plant community are
insensitive to to changes in precipitation regime. Such insensitivity
can buffer species against precipitation variability in this semi-arid
ecosystem, making them successful in the long run. Longer, chronic
precipitation alteration might reveal plant community shifts that we did
not observe \citep[e.g.,][]{Wilcox2016}. For example, a long-term
increase in water availability could allow species that do not bet hedge
to gain prominence and dominate the ecosystem functional response. Our
results suggest compositional shifts would have the largest impact at
high rainfall because the current community maintained consistent
ecosystem functional response at very low water availability.

@@

\begin{table}[ht]
\centering
\caption{Summary statistics from the posterior distributions of coefficients for each treatment.} 
\begingroup\normalsize
\begin{tabular}{llrrrr}
  \hline
Coefficient & Treatment & Posterior Mean & Posterior Median & Lower 95\% BCI & Upper 95\% BCI \\ 
  \hline
Intercept & Control & 0.16 & 0.14 & -1.22 & 1.69 \\ 
  Intercept & Drought & 0.13 & 0.05 & -1.69 & 2.39 \\ 
  Intercept & Irrigation & -0.24 & -0.18 & -2.11 & 1.53 \\ 
  Slope & Control & 0.94 & 0.86 & -0.51 & 2.69 \\ 
  Slope & Drought & 0.65 & 0.57 & -0.99 & 2.62 \\ 
  Slope & Irrigation & 0.72 & 0.65 & -0.54 & 2.26 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}\begin{table}[ht]
\centering
\caption{Results from statistical tests for clustering and dispersion of community composition among precipitation treatments. `adonis' tests whether treatments form unique clusters in multidimensial space; `betadisper' tests whether treatments have similar dispersion. For both tests, \emph{P} values greater than 0.05 indicate there is no support that the treatments differ.} 
\begingroup\normalsize
\begin{tabular}{rlrrrr}
  \hline
Year & Test & n & d.f. & \emph{F} & \emph{P} \\ 
  \hline
2011 & adonis &  21 &   2 & 1.02 & 0.44 \\ 
  2011 & betadisper &  21 &   2 & 2.23 & 0.14 \\ 
  2012 & adonis &  22 &   2 & 1.10 & 0.34 \\ 
  2012 & betadisper &  22 &   2 & 0.21 & 0.81 \\ 
  2013 & adonis &  22 &   2 & 1.23 & 0.16 \\ 
  2013 & betadisper &  22 &   2 & 0.28 & 0.76 \\ 
  2014 & adonis &  22 &   2 & 0.95 & 0.56 \\ 
  2014 & betadisper &  22 &   2 & 0.35 & 0.71 \\ 
  2015 & adonis &  21 &   2 & 1.05 & 0.37 \\ 
  2015 & betadisper &  21 &   2 & 3.01 & 0.07 \\ 
  2016 & adonis &  21 &   2 & 1.07 & 0.36 \\ 
  2016 & betadisper &  21 &   2 & 0.50 & 0.62 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}

>>>>

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, all estimates for the effect of
soil moisture on ANPP overlapped zero, indicating the relationship is
weak and uncertain regardless of treatment. There was also large spatial
variation in ANPP within-years, which contributes to the uncertainty of
the soil moisture effect. Plant community composition was remarkably
stable over the course of the experiment and did not differ among
treatments.

@@

We first averaged the observed soil moisture for each day (\emph{d}) and
plot (\emph{i}), \(x_{i,d}\).
\new{Then we standardized the average of each plot within a plot group ($g$) by subtracting the mean ambient soil moisture of the plot group for the day ($\overline{x}_{g,d} = \sum^{i=1}_G x_{i,d} / n$) from each plot's daily average and dividing by the standard deviation of the ambient soil moisture in that plot group ($\sigma_{g,d}$): $x^*_{i,d} = \frac{x_{i,d} - \overline{x}_{g,d}}{\sigma_{g,d}}$.}
We then found the difference between the standardized ambient soil
moisture and the standardized drought and irrigation soil moisture
within each plot group. These transformations ensured that the treatment
effects in each plot were appropriately scaled by the local ambient
conditions within each plot group.

@@

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS (Moderate Resolution Imaging
Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer)
bands for NDVI. We regressed the greenness index against the dry biomass
weight from the ten calibration plots to convert the greenness index to
ANPP. We fit regressions to a MODIS-based index and an AVHRR-based index
for each year and retained the regression with the better fit based on
\(R^2\) values. We then predicted ANPP using the best regression
equation for each year (Appendix 1).
\new{Our analysis, described below, can be done using NDVI directly, but we focus on ANPP to more easily relate our findings to previous work and because our results do not change if we analyze NDVI directly (Appendix 2).}

@@

Our results suggest the species in our focal plant community are
insensitive to to changes in precipitation regime. Such insensitivity
can buffer species against precipitation variability in this semi-arid
ecosystem, making them successful in the long run. Longer, chronic
precipitation alteration might reveal plant community shifts that we did
not observe \citep[e.g.,][]{Wilcox2016}. For example, a long-term
increase in water availability could allow species that do not bet hedge
to gain prominence and dominate the ecosystem functional response.
\new{Time \emph{per se} may be especially relevant in our focal ecosystem where the perennial species are long-lived, meaning compositional turnover may take many more years than we report on here.}
Our results suggest compositional shifts would have the largest impact
at high rainfall because the current community maintained consistent
ecosystem functional response at very low water availability.

@@

\begin{table}[ht]
\centering
\caption{Summary statistics from the posterior distributions of coefficients for each treatment.} 
\begingroup\normalsize
\begin{tabular}{llrrrr}
  \hline
Coefficient & Treatment & Posterior Mean & Posterior Median & Lower 95\% BCI & Upper 95\% BCI \\ 
  \hline
Intercept & Control & 0.16 & 0.14 & -1.22 & 1.69 \\ 
  Intercept & Drought & 0.13 & 0.05 & -1.69 & 2.39 \\ 
  Intercept & Irrigation & -0.24 & -0.18 & -2.11 & 1.53 \\ 
  Slope & Control & 0.94 & 0.86 & -0.51 & 2.69 \\ 
  Slope & Drought & 0.65 & 0.57 & -0.99 & 2.62 \\ 
  Slope & Irrigation & 0.72 & 0.65 & -0.54 & 2.26 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}\begin{table}[ht]
\centering
\caption{Results from statistical tests for clustering and dispersion of community composition among precipitation treatments. `adonis' tests whether treatments form unique clusters in multidimensial space; `betadisper' tests whether treatments have similar dispersion. For both tests, \emph{P} values greater than 0.05 indicate there is no support that the treatments differ.} 
\begingroup\normalsize
\begin{tabular}{rlrrrr}
  \hline
Year & Test & n & d.f. & \emph{F} & \emph{P} \\ 
  \hline
2011 & adonis &  21 &   2 & 1.02 & 0.43 \\ 
  2011 & betadisper &  21 &   2 & 2.22 & 0.14 \\ 
  2012 & adonis &  22 &   2 & 1.09 & 0.28 \\ 
  2012 & betadisper &  22 &   2 & 0.21 & 0.81 \\ 
  2013 & adonis &  22 &   2 & 1.25 & 0.16 \\ 
  2013 & betadisper &  22 &   2 & 0.44 & 0.65 \\ 
  2014 & adonis &  22 &   2 & 0.95 & 0.55 \\ 
  2014 & betadisper &  22 &   2 & 0.35 & 0.71 \\ 
  2015 & adonis &  21 &   2 & 1.05 & 0.39 \\ 
  2015 & betadisper &  21 &   2 & 3.01 & 0.07 \\ 
  2016 & adonis &  21 &   2 & 1.07 & 0.33 \\ 
  2016 & betadisper &  21 &   2 & 0.49 & 0.62 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}
